---
title: "Random Effects"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Installation

```{r, eval = FALSE}
install.packages("pak")                   # development version
pak::pak("gcol33/tulpaRatio")
```

## Overview

This vignette covers random effects in numdenom, including random intercepts, random slopes, crossed effects, and nested effects. numdenom uses lme4-style formula syntax with extensions for correlated vs. uncorrelated slopes.

# Random Intercepts

## Single Grouping Factor

```{r}
# Random intercept for site
fit <- tratio(
  count | effort ~ depth + (1 | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This estimates a site-specific intercept $b_{\text{site}[i]}$ for each observation $i$:

$$
\log(\mu_i) = \beta_0 + \beta_1 \cdot \text{depth}_i + b_{\text{site}[i]}
$$

where $b_j \sim \text{Normal}(0, \sigma_{\text{site}}^2)$.

## Multiple (Crossed) Grouping Factors

When observations are classified by multiple independent grouping factors:

```{r}
# Crossed random effects: site and year
fit <- tratio(
  count | effort ~ depth + (1 | site) + (1 | year),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This estimates:
- $b^{\text{site}}_j \sim \text{Normal}(0, \sigma_{\text{site}}^2)$
- $b^{\text{year}}_k \sim \text{Normal}(0, \sigma_{\text{year}}^2)$

The linear predictor becomes:

$$
\log(\mu_i) = \beta_0 + \beta_1 \cdot \text{depth}_i + b^{\text{site}}_{\text{site}[i]} + b^{\text{year}}_{\text{year}[i]}
$$

# Nested Random Effects

When groups are hierarchically nested (plots within sites):

```{r}
# Nested: plots within sites
fit <- tratio(
  count | effort ~ depth + (1 | site/plot),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This expands to:

```{r}
# Equivalent explicit form
fit <- tratio(
  count | effort ~ depth + (1 | site) + (1 | site:plot),
  data = df,
  family = ratiod_poisson_gamma()
)
```

The `site:plot` interaction creates unique identifiers for each plot-within-site combination.

### Deep Nesting

For deeper hierarchies:

```{r}
# region/site/plot expands to:
# (1 | region) + (1 | region:site) + (1 | region:site:plot)
fit <- tratio(
  count | effort ~ x + (1 | region/site/plot),
  data = df,
  family = ratiod_poisson_gamma()
)
```

# Random Slopes

Random slopes allow the effect of a predictor to vary by group.

## Correlated Intercept and Slope

Use single `|` for correlated random effects:

```{r}
# Intercept and depth effect vary by site, with correlation
fit <- tratio(
  count | effort ~ depth + (1 + depth | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This estimates:

$$
\begin{pmatrix} b_{0j} \\ b_{1j} \end{pmatrix} \sim \text{MVN}\left(\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} \sigma_0^2 & \rho\sigma_0\sigma_1 \\ \rho\sigma_0\sigma_1 & \sigma_1^2 \end{pmatrix}\right)
$$

The correlation $\rho$ captures whether sites with higher intercepts also tend to have steeper slopes.

## Uncorrelated Intercept and Slope

Use double `||` for independent random effects:

```{r}
# Intercept and depth effect vary by site, independently
fit <- tratio(
  count | effort ~ depth + (1 + depth || site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This is computationally simpler and appropriate when you don't expect correlation between intercepts and slopes.

## Multiple Slopes

```{r}
# Multiple correlated slopes
fit <- tratio(
  count | effort ~ depth + temp + (1 + depth + temp | site),
  data = df,
  family = ratiod_poisson_gamma()
)

# Multiple uncorrelated slopes
fit <- tratio(
  count | effort ~ depth + temp + (1 + depth + temp || site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

# Formula Expansion

numdenom supports lme4-style formula expansion for random slopes.

## Interactions

```{r}
# x*z expands to x + z + x:z
fit <- tratio(
  count | effort ~ x*z + (1 + x*z | site),
  data = df,
  family = ratiod_poisson_gamma()
)
# Random slopes for x, z, and their interaction
```

## Polynomials

```{r}
# Quadratic random slope
fit <- tratio(
  count | effort ~ x + I(x^2) + (1 + x + I(x^2) | site),
  data = df,
  family = ratiod_poisson_gamma()
)

# Or using poly()
fit <- tratio(
  count | effort ~ poly(x, 2) + (1 + poly(x, 2) | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Parameter Naming

numdenom cleans parameter names for readability:

| Formula Term | Parameter Name |
|--------------|----------------|
| `x:z` | `x_z` |
| `I(x^2)` | `x_pow2` |
| `poly(x, 2)1` | `x_poly1` |
| `poly(x, 2)2` | `x_poly2` |
| `log(x)` | `log_x` |
| `scale(x)` | `x_scaled` |

# Shared Random Effects

By default, numdenom shares random effects between numerator and denominator when the same grouping factor appears in both. This is the core of the numdenom philosophy (see `vignette("philosophy")`).

## Default: Shared Structure

```{r}
# Site effect shared between catch and effort processes
fit <- tratio(
  catch | effort ~ depth + (1 | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

The site effect enters both linear predictors identically:

$$
\log(\mu_{\text{catch}}) = \beta_0 + \beta_1 \cdot \text{depth} + \mathbf{b_{\text{site}}}
$$
$$
\log(\mu_{\text{effort}}) = \alpha_0 + \mathbf{b_{\text{site}}}
$$

Shared effects cancel in the ratio, removing spurious variation from common unmeasured drivers.

## Explicit Shared Structure

```{r}
# Only site is shared, observer is process-specific
fit <- tratio(
  catch | effort ~ (1 | site) + (1 | observer),
  shared = ~ (1 | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Independent Structure (Use with Caution)

```{r}
# Request independence (triggers warning)
fit <- tratio(
  successes | trials ~ treatment + (1 | block),
  shared = ~ 0,
  data = df,
  family = ratiod_binomial()
)
```

# Extracting Random Effects

```{r}
# Posterior summaries for random effects
re <- extract_ranef(fit)

# By term
re$site       # Site-level effects
re$site_plot  # Nested plot effects

# For random slopes
re$site$intercept  # Intercept component
re$site$depth      # Slope component
```

# Prior Specification

## Default PC Priors

numdenom uses penalized complexity (PC) priors that favor simpler models:

```{r}
# Default: P(sigma > 1) = 0.01
fit <- tratio(..., priors = ratiod_priors())
```

## Custom Priors

```{r}
# Expect smaller random effects
fit <- tratio(
  ...,
  priors = ratiod_priors(
    sigma_re_scale = 0.5  # Prior scale for RE standard deviations
  )
)
```

# Latent Factors for Unmeasured Confounders

When you suspect unmeasured confounders are driving both numerator and denominator but don't have a natural grouping structure, latent factors provide an alternative to random effects.

## Single Latent Factor

```{r}
# Capture shared unmeasured variation
fit <- tratio(
  count | effort ~ depth + (1 | site),
  data = df,
  family = ratiod_poisson_gamma(),
  latent = latent_factor(n_factors = 1)
)
```

Latent factors are observation-level random effects that enter both linear predictors:

$$
\log(\mu_{\text{count}}) = \beta_0 + \beta_1 \cdot \text{depth} + b_{\text{site}} + f_i \cdot \sigma_f
$$
$$
\log(\mu_{\text{effort}}) = \alpha_0 + b_{\text{site}} + f_i \cdot \sigma_f
$$

where $f_i \sim \text{Normal}(0, 1)$ with sum-to-zero constraint.

## Multiple Latent Factors

```{r}
# Multiple independent unmeasured confounders
fit <- tratio(
  count | effort ~ depth,
  data = df,
  family = ratiod_poisson_gamma(),
  latent = latent_factor(n_factors = 2)
)
```

## Latent Factors vs Random Effects

| Feature | Random Effects | Latent Factors |
|---------|---------------|----------------|
| Grouping | Required (site, year, etc.) | None needed |
| Level | Group-level | Observation-level |
| Use case | Known grouping structure | Residual confounding |
| Dimensionality | 1 per group | K factors × N observations |

Use latent factors when:
- You suspect unmeasured confounders but don't know the grouping structure
- Residual variation after random effects is still correlated
- You want to capture observation-level shared structure

## Prior Specification

```{r}
# Stronger regularization on latent factors
fit <- tratio(
  count | effort ~ depth,
  data = df,
  family = ratiod_poisson_gamma(),
  latent = latent_factor(
    n_factors = 1,
    prior = prior_pc(U = 0.5, alpha = 0.01)  # P(sigma > 0.5) = 0.01
  )
)
```

## See Also

- `vignette("getting-started")` - Basic usage
- `vignette("philosophy")` - Why shared structure matters
- `vignette("spatial-temporal")` - Spatial and temporal extensions
