title: "Spatial and Temporal Models"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial and Temporal Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Installation

```{r, eval = FALSE}
install.packages("pak")                   # development version
pak::pak("gcol33/tulpaRatio")
```

## Overview

This vignette covers spatial and temporal extensions in numdenom. These allow borrowing strength across neighboring locations or time points, improving estimates in data-sparse regions.


# Spatial Models

numdenom supports two spatial priors for areal (discrete region) data:

| Function | Prior | Use Case |
|----------|-------|----------|
| `spatial_car()` | Intrinsic CAR (ICAR) | Simple spatial smoothing |
| `spatial_bym2()` | BYM2 | Interpretable spatial fraction |

Both require an adjacency matrix defining which regions are neighbors.

## Creating Adjacency Matrices

### From Coordinates (Grid Data)

```{r}
# Camera trap grid
n <- 25
coords <- expand.grid(x = 1:5, y = 1:5)

# Queen contiguity (8 neighbors)
adj <- matrix(0, n, n)
for (i in 1:n) {
  for (j in 1:n) {
    if (i != j) {
      dist <- sqrt((coords$x[i] - coords$x[j])^2 +
                   (coords$y[i] - coords$y[j])^2)
      if (dist <= sqrt(2) + 0.01) adj[i, j] <- 1
    }
  }
}
```

### From sf Objects

```{r}
library(sf)
library(spdep)

# From shapefile
regions <- st_read("regions.shp")
nb <- poly2nb(regions, queen = TRUE)
adj <- nb2mat(nb, style = "B", zero.policy = TRUE)
```

## ICAR Model: `spatial_car()`

The intrinsic conditional autoregressive (ICAR) prior encourages neighboring regions to have similar effects:

$$
\phi_i | \phi_{-i} \sim \text{Normal}\left(\frac{1}{n_i}\sum_{j \sim i} \phi_j, \frac{\sigma^2}{n_i}\right)
$$

where $j \sim i$ denotes neighbors of region $i$.

```{r}
fit <- tratio(
  count | effort ~ x + (1 | region),
  spatial = spatial_car(
    adj = adj,
    level = "group",
    group_var = "region"
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Parameters

- `adj`: Square binary adjacency matrix
- `level`: `"group"` (one effect per group) or `"obs"` (one per observation)
- `group_var`: Variable name matching the grouping factor (required if `level = "group"`)

## BYM2 Model: `spatial_bym2()`

The BYM2 reparameterization (Riebler et al., 2016) separates spatial structure from unstructured heterogeneity:

$$
\theta_i = \sigma \left(\sqrt{\rho} \cdot \phi_i^* + \sqrt{1-\rho} \cdot v_i\right)
$$

where:
- $\phi_i^*$ is the scaled ICAR component
- $v_i \sim \text{Normal}(0, 1)$ is unstructured noise
- $\rho \in [0, 1]$ is the proportion of variance that is spatial

```{r}
fit <- tratio(
  cases | population ~ age + (1 | county),
  spatial = spatial_bym2(
    adj = adj,
    level = "group",
    group_var = "county"
  ),
  data = df,
  family = ratiod_binomial()
)
```

The $\rho$ parameter is directly interpretable: $\rho = 0.8$ means 80% of the random effect variance is spatially structured.

## Extracting Spatial Effects

```{r}
# Get posterior summaries
spatial_effects <- extract_spatial(fit)

# Map visualization
library(ggplot2)
library(sf)

regions$spatial_effect <- spatial_effects$mean
ggplot(regions) +
geom_sf(aes(fill = spatial_effect)) +
  scale_fill_gradient2()
```


# Temporal Models

numdenom supports temporal priors for time series structure:

| Function | Prior | Use Case |
|----------|-------|----------|
| `temporal_rw()` | Random walk | Smooth trends |
| `temporal_ar1()` | AR(1) | Mean-reverting dynamics |

## Random Walk: `temporal_rw()`

First-order random walk (RW1):

$$
\gamma_t | \gamma_{t-1} \sim \text{Normal}(\gamma_{t-1}, \sigma_\gamma^2)
$$

Second-order random walk (RW2) for smoother trends:

$$
\gamma_t | \gamma_{t-1}, \gamma_{t-2} \sim \text{Normal}(2\gamma_{t-1} - \gamma_{t-2}, \sigma_\gamma^2)
$$

```{r}
# First-order random walk
fit <- tratio(
  count | effort ~ x + (1 | site),
  temporal = temporal_rw(
    time_var = "year",
    order = 1
  ),
  data = df,
  family = ratiod_poisson_gamma()
)

# Second-order for smoother trends
fit <- tratio(
  count | effort ~ x,
  temporal = temporal_rw(
    time_var = "year",
    order = 2
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Cyclic Random Walk

For seasonal patterns that wrap around (e.g., month 12 neighbors month 1):

```{r}
fit <- tratio(
  count | effort ~ x,
  temporal = temporal_rw(
    time_var = "month",
    order = 1,
    cyclic = TRUE
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## AR(1): `temporal_ar1()`

Autoregressive order 1:

$$
\gamma_t = \rho \cdot \gamma_{t-1} + \epsilon_t, \quad \epsilon_t \sim \text{Normal}(0, \sigma_\gamma^2)
$$

The correlation parameter $\rho \in (-1, 1)$ controls persistence.

```{r}
fit <- tratio(
  count | effort ~ x,
  temporal = temporal_ar1(
    time_var = "year"
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Group-Specific Temporal Effects

By default, temporal effects are shared across all groups. For group-specific trends:

```{r}
fit <- tratio(
  count | effort ~ (1 | site),
  temporal = temporal_rw(
    time_var = "year",
    group_var = "site",
    shared = FALSE
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Extracting Temporal Effects

```{r}
# Year-level effects
temporal_effects <- extract_temporal(fit)

# Plot trend
library(ggplot2)
ggplot(temporal_effects, aes(x = year, y = mean)) +
  geom_ribbon(aes(ymin = q2.5, ymax = q97.5), alpha = 0.2) +
  geom_line() +
  labs(y = "Temporal effect")
```


# Combining Spatial and Temporal

For spatiotemporal data, include both:

```{r}
fit <- tratio(
  count | effort ~ x + (1 | region),
  spatial = spatial_bym2(adj, level = "group", group_var = "region"),
  temporal = temporal_rw(time_var = "year"),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This fits additive spatial and temporal effects. For interaction effects, see the next section.


# Spatiotemporal Interaction

When spatial patterns change over time (or temporal trends differ across space), additive models may be insufficient. numdenom supports spatiotemporal interaction effects following Knorr-Held (2000).

## Interaction Types

| Type | Description | Parameters |
|------|-------------|------------|
| Type I | Unstructured (IID) | $S \times T$ |
| Type II | Structured time at each location | $S \times T$ |
| Type III | Structured space at each time | $S \times T$ |
| Type IV | Fully structured (Kronecker) | $S \times T$ |
| Separable | For GP spatial/temporal | $S \times T$ |

### Type I: Unstructured Interaction

The simplest form assumes independent random effects for each space-time combination:

$$
\delta_{st} \stackrel{iid}{\sim} \text{Normal}(0, \sigma_\delta^2)
$$

```{r}
fit <- tratio(
  count | effort ~ x,
  data = df,
  family = ratiod_poisson_gamma(),
  spatiotemporal = spatiotemporal(
    spatial = spatial_car(adj, level = "group", group_var = "region"),
    temporal = temporal_rw1("year"),
    type = "I"
  )
)
```

### Type II: Structured Time at Each Location

Each location has its own temporal random walk:

$$
\delta_{\cdot t}^{(s)} \sim \text{RW}(\sigma^2)
$$

This captures location-specific temporal trends.

```{r}
fit <- tratio(
  count | effort ~ x,
  data = df,
  family = ratiod_poisson_gamma(),
  spatiotemporal = spatiotemporal(
    spatial = spatial_car(adj, level = "group", group_var = "region"),
    temporal = temporal_rw1("year"),
    type = "II"
  )
)
```

### Type III: Structured Space at Each Time Point

Each time point has its own spatial field:

$$
\delta_{s \cdot}^{(t)} \sim \text{ICAR}(\tau)
$$

This captures time-specific spatial patterns.

```{r}
fit <- tratio(
  count | effort ~ x,
  data = df,
  family = ratiod_poisson_gamma(),
  spatiotemporal = spatiotemporal(
    spatial = spatial_car(adj, level = "group", group_var = "region"),
    temporal = temporal_rw1("year"),
    type = "III"
  )
)
```

### Type IV: Fully Structured (Kronecker)

Precision is the Kronecker product of spatial and temporal precision:

$$
Q_\delta = Q_s \otimes Q_t
$$

This is the most constrained and often most appropriate model.

```{r}
fit <- tratio(
  count | effort ~ x,
  data = df,
  family = ratiod_poisson_gamma(),
  spatiotemporal = spatiotemporal(
    spatial = spatial_car(adj, level = "group", group_var = "region"),
    temporal = temporal_rw1("year"),
    type = "IV"
  )
)
```

## Non-Separable Spatiotemporal GP

For continuous space-time data, `type = "nonsep_gp"` puts a covariance on the
space-time grid rather than a GMRF stencil on it. The interaction carries its
own `coords`, so an areal main effect pairs with a continuous interaction over
the same units:

```{r}
fit <- tratio(
  count | effort ~ x,
  data = df,
  family = ratiod_poisson_gamma(),
  spatiotemporal = spatiotemporal(
    spatial = spatial_car(adj, level = "group", group_var = "region"),
    temporal = temporal_rw1("date"),
    type = "nonsep_gp",
    nonsep_type = "gneiting",
    coords = ~ lon + lat
  )
)
```

Non-separability types:

- `"product"`: the separable covariance, `C_s(h) C_t(u)`. `type = "separable"`
  is this arm.
- `"gneiting"`: Gneiting (2002) class

An additive covariance `C_s(h) + C_t(u)` is not among them. On a complete
`S x T` grid its rank is `S + T - 1`, and the directions it drops are exactly
the space-time interactions the block exists to carry.

Both ranges are estimated on their own log scale, alongside the interaction's
precision. The field is approximated by an NNGP over the grid, with `nn`
neighbours per cell chosen on a joint space-time metric.

## Extracting Spatiotemporal Effects

```{r}
# Summary format
st_effects <- spatiotemporal_effects(fit, format = "summary")
head(st_effects)

# Array format (S x T x draws)
st_array <- spatiotemporal_effects(fit, format = "array")

# Visualization
plot(st_effects, type = "heatmap")
plot(st_effects, type = "time_series")
```

## Model Selection

Compare interaction types using LOO-CV:

```{r}
fit_type1 <- tratio(..., spatiotemporal = spatiotemporal(..., type = "I"))
fit_type4 <- tratio(..., spatiotemporal = spatiotemporal(..., type = "IV"))

loo_compare(loo(fit_type1), loo(fit_type4))
```


## Prior Specification

### Spatial Priors

```{r}
fit <- tratio(
  ...,
  spatial = spatial_car(adj, level = "group", group_var = "region"),
  priors = ratiod_priors(
    tau_spatial_shape = 1.0,  # Gamma shape for precision
    tau_spatial_rate = 0.01   # Gamma rate for precision
  )
)
```

### Temporal Priors

```{r}
fit <- tratio(
  ...,
  temporal = temporal_rw(time_var = "year"),
  priors = ratiod_priors(
    tau_temporal_shape = 2.0,
    tau_temporal_rate = 0.5
  )
)
```


## See Also

- `vignette("getting-started")` - Basic usage
- `vignette("workflows")` - Complete analysis examples
- `?spatial_car`, `?spatial_bym2` - Function documentation
- `?temporal_rw`, `?temporal_ar1` - Function documentation
